Product plots

### 2011-03-30 prodplots-infovis.tex
Add a teaser banner
Tweaking banner
Lots more work

@@

@@

\abstract{We propose a new framework for visualising counts or proportions, including existing graphics such as bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams. We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.}

@@

% \teaser{
% \centering
% \includegraphics[width=16cm]{tx-cancer}
% \caption{One of these plots doesn't belong. These six plots show choropleth maps of cancer deaths in Texas, where darker colors = more deaths. Can you spot which of the six plots is made from a real dataset and not simulated under the null hypothesis of spatial independence?  If so, you've provided formal statistical evidence that deaths from cancer have spatial dependence. See Section~\ref{sec:solutions} for the answer.}
% \label{fig:teaser}
% }

@@

Tables of counts or proportions are an extremely common form of data. Correspondingly, there are many visualisations to display them. In this paper, we develop a framework that encompasses many existing visualisations, from bar charts to treemaps to pie charts, showing how graphics that previously seemed unrelated in fact share a deep underlying connection, and providing a systematic foundation for the developing new types of count visualisations \citep{cox:1978}.

Our framework focusses on area charts, where the area of a graphical element is proportional to the underlying count. We call our framework product plots in allusion to two products: the product of width and height to compute area, and the product of conditional and marginal distributions to produce joint distributions. A key development of our framework is the inverse operation: the factorisation of area plots. This allows easy extension from low-dimensional graphical primitives to high-dimensional data sets.

Section~\ref{sec:primitives} we motivate the three constraints that give rise to the fundamental graphical component of productplots: the rectangle. The rectangle gives rises to three fundamental 1d atoms (bar, spine, tile) and one fundamental 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include even more visualisations including histograms, pie charts, cascaded treemaps and weighted plots.

Section~\ref{sec:combination} provides the mathematical framework that extends the simple 1d and 2d atoms to display data of any dimensionality, by showing just as any high dimensional joint distribution can be represented as a product of low-dimensional conditional and marginal distributions, any area plot can be decomposed into products of simple low-d primitives. Using this framework, Section~\ref{sec:existing} shows how 13 existing named graphics are just special cases of this general pattern.

To illustrate these ideas, we will use the same data for all examples. The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}. The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976. We combine data for 25 years to gives 51\,020 observations, and of the over 5\,000 variables, we select eight related to happiness, as described in Table~\ref{tbl:happy}.

\begin{table*}[htb]
  \centering
  \begin{tabular}{llp{9cm}}
    \toprule
    Variable & Description & Values \\
    \midrule
    {\sf age} & age in years & 18--89 \\
    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
    {\sf finrela} & relative financial status & far above, above average, average, below average, far below \\
    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
    {\sf health} & health & excellent, good, fair, poor \\
    {\sf marital} & marital status & \key{married} married, \key{never-married} never married, \key{divorced} divorced, \key{widowed} widowed, \key{separated} separated \\
    {\sf sex} & sex & \key{female} female, \key{male} male\\
    {\sf wtsall} & probability weight & 0.43--6.42 \\
    {\sf year} & year of survey & 1972--2006 \\
    \bottomrule
  \end{tabular}
  \caption{Description of sample data. Common plot colours are shown next to respective levels.}
  \label{tbl:happy}
\end{table*}

@@

All area plots correspond to tilings of the plane. We could consider partitions of higher-d spaces (e.g. 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d. There are many possible ways to tile the plane. To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:

@@

  \item Area must be {\bf proportional} to count. Because the total area for a graphic is usually constrained, and area plot partitions the total, area plots display proportions, not counts. From here on, without loss of generality, we will refer to visualisation of proportions, rather than counts.

  \item Partitions must be {\bf rectangular}, because rectangles are easier to compare than arbitrary polygons \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive in the sense that we can always tile a rectangle with smaller rectangles. This is not true for most shapes.

  \item Partitions must be {\bf disjoint}. To be able to see the complete area, each rectangle must be non-overlapping. This does not imply that the tiling must be space-filling.

@@

\noindent Each of three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.

These constraints give rise to four graphical primitives. Section~\ref{sub:part-1d} describes the partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.

% Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$. Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively. Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$. This is useful because it is only necessary to inspect one border to compare magnitudes.

@@

1d primitives display one dimensional proportion data, i.e.\ proportions broken down by a single variable. There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}.

@@

  \item {\bf spine}s: width is proportional to value, height occupies full range. Spines are space filling and can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'') by splitting the largest dimension. The name spine is evocative of books sitting on a library shelf \citep{hummel:1996}. Bars and spines are identical when visualising uniformly distributed data.

  \item {\bf tile}s: no restrictions on height or width, just tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1. This is partitioning used by the squarified treemap \citep{bruls:1999}.

@@

Each of these three displays has different strengths and weaknesses. The perceptual task associated with bar charts is the simplest, but they do not occupy all the area. Spines and tiles are harder to read \citep{heer:2010}, but occupy the complete space and so work better recursively.

@@

2d primitives display two dimensional proportion data, i.e.\ proportions broken down by two variables. We are currently aware of one primitive for 2d data, the {\bf fluct}, derived from the fluctuation diagram \citep{hofmann:2000}. The fluct has height and width proportional to the square root of the proportion. Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons in two directions.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9 \linewidth]{part-fluct}
    \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
  \caption{(Left) A fluctuation diagram showing distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

@@

@@

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful. One way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}. A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs. A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all value must sum to one.

\subsection{Products of probabilities}

Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness. The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex. These correspond to restriction the row or column sums to equal one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''. Finally, the last table display the two {\bf marginal distributions} of sex and happiness. These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.

@@

  \caption{The distribution of happiness and sex, displayed in three equivalent ways.  (Left) Joint distribution. Overall table sums to one. (Middle) Conditional distribution of sex given happiness and marginal distribution of happiness. (Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
  \label{fig:2d-table}
\end{figure}

Formally, the conditional distribution is written as $f(x | y)$ and is equal to $f(x, y) / f(y)$. This definition illustrates that given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y) f(y)$ \footnotemark. This definition extends in a straightforward way to higher dimensions. For example, a 3d pmf can be written as the product of 2d and 1d conditional and marginal distributions as follows:

\footnotetext{Because the parameters of a pmf are identifying, statisticians often fail to distinguish the different functions.  To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{(X|Y=y)}(x | y) f_Y(y)$}

\begin{itemize}
  \setlength{\itemsep}{0em}
  \item $f(x, y, z) = f(z) f(x, y | z)$
  \item $f(x, y, z) = f(y, z) f(x | y, z) $
  \item $f(x, y, z) = f(z) f(y | z) f(x | y, z)$
\end{itemize}

This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s products. It remains to find an graphical analog of this multiplication.

\subsection{Area is a product of height and width}

We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width. It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple one primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart. From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$. The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.3\linewidth]{fact-happy} $\times$ %
  \includegraphics[width=0.3\linewidth]{fact-happy|sex} = %
  \includegraphics[width=0.3\linewidth]{fact-happy-sex}

  \includegraphics[width=0.3\linewidth]{fact-happy-2} $\times$ %
  \includegraphics[width=0.3\linewidth]{fact-happy|sex-2} = %
  \includegraphics[width=0.3\linewidth]{fact-happy-sex-2}

@@

Figure \ref{fig:recursive} shows the steps along by which a 3d dataset is displayed.  We first display $f(martial status)$, then $f(marital sex, sex) = f(sex | marital status) f(marital status)$, and finally $f(marital sex, sex, happiness) = f(happiness | sex, marital status) f(sex | marital status) f(marital status)$.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-comb-1}%
    \includegraphics[width=0.33\linewidth]{part-comb-2}%
    \includegraphics[width=0.33\linewidth]{part-comb-3}
  \caption{Conditional on marital status, are men or women happier?  This figure shows the construction of $f(happy, sex, marital)$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.  For all levels of marital status, men are slightly less happy.}
  \label{fig:recursive}
\end{figure}

% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
% 
% Writing the full pmf is tedious, so we define an algebra based on the Wilkinson's modelling specification \citep{wilkinson:1973}:
% 
% \begin{itemize}
%   \setlength{\itemsep}{0em}
%   \item \verb|~ a| represents $f(a)$,
%   \item \verb|~ a + b| represents $f(a, b)$
%   \item \verb!~ a | b! represents $f(a | b)$
%   \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
% \end{itemize}
%

% To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives. That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.

Different combinations of partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine. These displays show the same data but are arranged in such a way to make different comparisons easier. The second plot, with bars nested inside spines, also illustrates important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level. The bars are scaled to be as high as possible, without overflowing the bounding area.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because the perceptual task is simpler.  Depending on the comparison we are most interested in, we can make it easier to compare across wealth for fixed health, or health for fixed wealth.

For this data, for a fixed income level, happiness always increases as health increases. The same is not true for fixed health: those with poor or fair health are actually less happy if they are far above average, compared to just above average.

\begin{figure*}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-fluctuation}%
    \includegraphics[width=0.33\linewidth]{part-equal-area}%
    \includegraphics[width=0.33\linewidth]{part-equal-area-2}
  \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.  (Middle) $f(happy | health, finrela)$. We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness. Healthier and richer people are happier: maybe money does buy happiness? (Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine. (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
  \label{fig:part-cond}
\end{figure*}

@@

Many existing plots fall into this framework. The primitives already have their own names:

@@

  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.

  \item {\bf Nested} bar chart \citep{peltier:2009} (2d).  2 hbars.

  \item {\bf Equal bin size} \citep{hofmann:2000} plot (3d): fluct and vspine, conditioned on the first two variables.

@@

  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and 1 vspine.

  \item {\bf Treemap} \citep{shneiderman:1992} (nd): n spines.

  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): n tiles.

@@

Trellis graphics \citep{becker:1996}, also known as lattice, facetted and conditioned graphics, are another related display. They use categorical variables to generate multiple panels, each containing a plot of the subset of the data. Trellised area plots also fall into our framework and can be created by conditioning on the trellising variables.

@@

On top of this basic framework, it is useful to consider a few variations and extensions. Instead of count of discrete levels, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.

@@

We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is subtly different, and suggests that we don't need to worry about weights for this plot. In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities. Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{wt-count}%
    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights.}
  \label{fig:weighting}
\end{figure}

>>>>

% TO DO:
% * look up Heike's kronecker product notation
% * complete display considerations
% * complete conclusions and plans for future work

@@

\newcommand{\includegraphicsinline}[2]{\begin{minipage}{#1}\includegraphics[width=\linewidth]{#2}\end{minipage}}

@@

\abstract{

We propose a new framework for visualising tables counts, proportions or probabilities. We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions. The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.

}

@@

\teaser{
\centering
\includegraphics[width=0.16\linewidth]{banner-1}%
\includegraphics[width=0.16\linewidth]{banner-2}%
\includegraphics[width=0.16\linewidth]{banner-5}%
\includegraphics[width=0.16\linewidth]{banner-6}%
\includegraphics[width=0.16\linewidth]{banner-3}%
\includegraphics[width=0.16\linewidth]{banner-4}
\caption{A selection of plots encompassed by the product plots framework.  All display the same data (the distribution of happiness and marital status), but each supports a different question. From left to right: mosaic plot, treemap, stacked  bar chart, fluctuation diagram, and two new plots that don't have names. In all plots, area is proportional to probability. }
\label{fig:teaser}
}

@@

Tables of counts, proportions and probabilities are an extremely common form of data, and many researchers have developed visualisations to display them. In this paper, we develop a framework that encompasses many existing visualisations, from bar charts to treemaps to pie charts, showing how graphics that previously seemed unrelated in fact share a deep underlying connection \citep{cox:1978}. Our framework makes it easier to describe and create visualisations of categorical data, shows how existing methods are related and where new methods can be developed, and will make it easier to match questions about categorical data to the visualisations that will provide revealing answers.

Our framework focusses on area charts, where the area of a graphical element is proportional to the underlying count, proportion, or probability. We call our framework \emph{product plots} in allusion to two products: the product of width and height to generate area, and the product of conditional and marginal distributions to produce joint distributions. A key development of the products plot framework is the inverse operation: the factorisation of high-dimensional data to products of low-dimensional plots. This allows us to  combine simple, low-dimensional graphical primitives to display complex, high-dimensional data.

We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots, the rectangle. We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.

Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality. This framework is based on the fundamental statistical idea that high any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions. Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.

Section~\ref{sec:display} discusses some considerations for the display of product plots, and Section~\ref{sec:package} introduces the R package {\tt prodplot}, our reference implementation of the product plots framework. Finally, Section~\ref{sec:conclusion} discusses our plans for the future.

To illustrate these ideas, we will use the same data in all examples. The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}. The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976. We combine data for 25 years to yield 51\,020 observations, and of the over 5\,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.

\begin{table*}[htb]
  \centering
  \begin{tabular}{llp{9cm}}
    \toprule
    Variable & Description & Values \\
    \midrule
    {\sf age} & age in years & 18--89 \\
    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
    {\sf finrela} & relative financial status & far above, above average, average, below average, far below \\
    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
    {\sf health} & health & excellent, good, fair, poor \\
    {\sf marital} & marital status & \key{married} married, \key{never-married} never married, \key{divorced} divorced, \key{widowed} widowed, \key{separated} separated \\
    {\sf sex} & sex & \key{female} female, \key{male} male\\
    {\sf wtsall} & probability weight & 0.43--6.42 \\
    {\sf year} & year of survey & 1972--2006 \\
    \bottomrule
  \end{tabular}
  \caption{Description of sample data. Common plot colours are shown next to respective levels.  Colours from ColorBrewer palettes \citep{brewer:1994}.}
  \label{tbl:happy}
\end{table*}

@@

All area plots correspond to tilings of the (2d) plane. We could consider partitions of higher-d spaces (e.g. 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.

There are many possible ways to tile the plane. To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:

@@

  \item Area must be {\bf proportional} to count. This is the key constraint underlying all area plots. The total area for a graphic is usually constrained, which means that area plots display typically proportions or probabilities, rather than counts. We will use count, probability and proportion interchangeably throughout the paper.

  \item Partitions must be {\bf rectangular}, because the areas of rectangles are easier to compare than areas of other shapes \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive, in the sense that we can always tile a rectangle with smaller rectangles. This is not true for most shapes.

  \item Partitions must be {\bf disjoint}. To be able to see the complete area, each rectangle must be non-overlapping. Note that this does not imply that the tiling must be space-filling, and two of the four primitives, described next, are not.

@@

\noindent Each of the three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.

These constraints give rise to four graphical primitives. Section~\ref{sub:part-1d} describes the three partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.

@@

1d primitives display 1d \emph{data}, i.e.\ counts broken down by a single variable. There are three 1d primitives, as shown in Figure~\ref{fig:part-1d} and described below.

@@

  \item {\bf spine}s: width is proportional to value, height occupies full range. Spines are space filling and can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'') by splitting the largest dimension. The name spine is evocative of books sitting on a library shelf \citep{hummel:1996}. Bars and spines are indistinguishable when the underlying data is uniformly distributed.

  \item {\bf tile}s: no restrictions on height or width, just tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1. This is partitioning defined by the squarified treemap \citep{bruls:1999}.

@@

Each of these three displays has different strengths and weaknesses. It is easiest to compare the value associated with bars because the perceptual task is the easiest: comparing position on a common scale. Reading spines and tiles are harder because we must compare lengths or areas, but they occupy the complete space and so work better recursively.

@@

2d primitives display 2d \emph{data}, i.e.\ a count broken down by two variables. We are currently aware of only one primitive for 2d data, the {\bf fluct}, derived from the fluctuation diagram \citep{hofmann:2000}. The fluct has height and width proportional to the square root of the count. Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons both vertically and horizontally.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9 \linewidth]{part-fluct}
    \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
  \caption{(Left) A fluctuation diagram showing distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Thee three empty columns show that health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

@@

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}. A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs. A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.

\subsection{Joint distributions are the product of marginals and conditionals}

Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness. The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex. These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''. Finally, the last table displays the two {\bf marginal distributions} of sex and happiness. These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.

@@

  \caption{The distribution of happiness and sex, displayed in three equivalent ways.  (Top) Joint distribution. Overall table sums to one. (Middle) Conditional distribution of sex given happiness and marginal distribution of happiness. (Bottom) Conditional distribution of happiness given sex and marginal distribution of sex.}
  \label{fig:2d-table}
\end{figure}

Formally, a conditional distribution function is written $f(x | y)$ and is equal to $f(x, y) / f(y)$. This definition illustrates that an important statistical fact: given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y)\,f(y)$ \footnotemark. This definition extends in a straightforward way to higher dimensions. For example, a 3d joint distribution can be written as the product of 2d and 1d conditional and marginal distributions in the following three ways:

\begin{itemize}
  \setlength{\itemsep}{0em}
  \item $f(x, y, z) = f(z)\,f(x, y | z)$
  \item $f(x, y, z) = f(y, z)\,f(x | y, z) $
  \item $f(x, y, z) = f(z)\,f(y | z)\,f(x | y, z)$
\end{itemize}

\footnotetext{Because the parameters of a {\sc pmf} identify it, statisticians often fail to explicitly label the different functions. To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{X|Y=y}(x | y)\,f_Y(y)$}

This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation. It remains to find an graphical analog of this multiplication.

\subsection{Area is the product of height and width}

We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width. It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple 1d primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart. From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$. The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.

\begin{figure}[htbp]
  \centering
  \includegraphicsinline{0.3\linewidth}{fact-happy} = %
  \includegraphicsinline{0.3\linewidth}{fact-happy|sex} = %
  \includegraphicsinline{0.3\linewidth}{fact-happy-sex}

  \includegraphicsinline{0.3\linewidth}{fact-happy-2} $\times$ %
  \includegraphicsinline{0.3\linewidth}{fact-happy|sex-2} = %
  \includegraphicsinline{0.3\linewidth}{fact-happy-sex-2}

@@

Figure \ref{fig:recursive} shows a more complicated example: visualising a 3d distribution as a product of three low-dimensional distributions. We first display $f(marital)$, then $f(marital, sex) = f(sex | marital) f(marital)$, and finally $f(marital, sex, happy) = f(happy | sex, marital) f(sex | marital) f(marital)$. This plot uses two vspines and an hspine to produce a mosaic plot.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-comb-1}%
    \includegraphics[width=0.33\linewidth]{part-comb-2}%
    \includegraphics[width=0.33\linewidth]{part-comb-3}
  \caption{Conditional on marital status, are men or women happier?  This figure shows the construction of $f(happy, sex, marital)$ with (from left to right) a vspine by marital status, a vspine by marital status and sex, and a hspine by marital status, sex and happiness.  For all levels of marital status, men are slightly less happy.}
  \label{fig:recursive}
\end{figure}

Formally, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives. The following algorithm gives an rough idea of the computation:

\begin{itemize}
  \item Calculate the one marginal and $i$ conditional distributions.

  \item For each probability in the marginal distribution, divide the current bounds into $i$ sets of new bounds, one for each level of the categorical variable. The new bounds are found using the algorithm defined by the drawing primitive.

  \item For each new bound and matching conditional distribution, call the partition function recursively, dropping one element from the list of drawing primitives.

Different partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualising the joint distribution with a fluct, we could focus on the conditional distribution of marital status given age, or age given marital status. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine. These displays show the same data, but support different comparisons: on the left, we can see that most young people are unmarried, and on the right, that few unmarried people are over the age of 30. The right plot, with bars nested inside spines, also illustrates an important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level. The bars are scaled to be as tall as possible, without overflowing any bounding region.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because we must compare areas, not positions. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task. Depending on the comparison we are most interested in, we can make it easier to compare across wealth given health, or health given wealth. Here we see that for a fixed income level, better health is correlated to increased happiness. The same is not true for a fixed level of health: rich people with poor health seem to be less happy than poorer people in poor health.

\begin{figure*}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-fluctuation}%
    \includegraphics[width=0.33\linewidth]{part-equal-area}%
    \includegraphics[width=0.33\linewidth]{part-equal-area-2}
  \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.  Health is on the $x$-axis, financial status on the $y$-axis. (Middle) $f(happy | health, finrela)$. We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness. Healthier and richer people are happier: maybe money does buy happiness? (Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine. (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
  \label{fig:part-cond}
\end{figure*}

@@

Many existing plots fall into this framework. The low-d primitives already have their own names:

@@

  \item {\bf Stacked} bar chart (2d). $1$ hbar and $1$ vspine.

  \item {\bf Nested} bar chart \citep{peltier:2009} (2d). $2$ hbars.

  \item {\bf Equal bin size} plot \citep{hofmann:2000} (3d): $1$ fluct and $1$ vspine, conditioned on the first two variables.

@@

  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and $1$ vspine.

  \item {\bf Treemap} \citep{shneiderman:1992} (nd): $n$ spines.

  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): $n$ tiles.

@@

Trellis graphics \citep{becker:1996}, also known as latticed, facetted and conditioned graphics, are another related display. They use categorical variables to generate multiple panels, each containing a plot of the subset of the data. Trellised plots of area graphics also fall into our framework and can be created by conditioning on the trellising variables.

\section{Display considerations}
\label{sec:display}

Labelling product plots is challenging. In this paper, we use a carefully selected combination of colour and axis labels, as well displaying the process by which a product plot is created. Axis labels are only available under certain conditions: when we have rows or columns that all share the same dimensional index. This can occur because of the structure of the graphical primitive (e.g. bars, flucts), or because the display is of a conditional distribution. Take Figure~\ref{fig:marital} as an example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled. There are other ways to label the cells apart from colour. \citep{slingsby:2009} uses text labels, sized to occupy the space of the region they label. In the interactive setting, dynamic labelling in the form of tool tips is extremely helpful, as is the ability to use linked brushing to connect high-d plots to low-d plots.

We have observed that some aspect ratios are more aesthetically pleasing than others. This is particularly obvious for the fluctuation diagram where, in our experience, the plots are most appealing when the flucts are square. In other plots there does not seem to be an easy rule of thumb, but we wonder if aspect ratios close to the golden ratio might be more appealing. We are not aware of any previous work on aesthetics of aspect ratio, although they do seem to affect perception \citep{kong:2010}. More research in this area is needed, particularly to explore the balance of aesthetics and usability.

@@

On top of this basic framework, it is useful to consider a few variations and extensions. Instead of counts, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.

@@

We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is barely different, and suggests that we don't need to worry about weights for this plot. (Unfortunately for this dataset of we have been unable to find any plots where weighting makes a big difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities. Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{wt-count}%
    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights. In this case, there is very little difference between the plots.}
  \label{fig:weighting}
\end{figure}
